---
title: "Collapsing bindingDB compound-gene relationships"
output:
  html_document:
    theme: cosmo
    highlight: pygments
---

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(DT)
library(scales)
library(readr)

options(stringsAsFactors=FALSE)
```

```{r}
# Read bindingdb and remove non-human interactions
binding.db <- file.path('data', 'binding.tsv.gz') %>%
  readr::read_tsv() %>%
  dplyr::filter(organism == 'Homo sapiens') %>%
  dplyr::filter(! is.na(affinity_nM)) %>%
  dplyr::mutate(
    source=plyr::mapvalues(source, c('Curated from the literature by BindingDB'), c('BindingDB'))
  )

# View a subset of the data.frame
binding.db %>% dplyr::sample_n(200) %>% dplyr::select(-c(pubmed, doi)) %>% DT::datatable()
```


```{r}
# Read the drugbank to bindingDB fuzzy mappings produced using UniChem
# Restrict to compounds in drugbank
joined.df <- 'https://raw.githubusercontent.com/dhimmel/drugbank/3e87872db5fca5ac427ce27464ab945c0ceb4ec6/data/mapping/bindingdb.tsv' %>%
  readr::read_tsv() %>%
  dplyr::inner_join(binding.db)
```

`r nrow(joined.df)` compound--protein binding measurements are extracted for humans when restricting to DrugBank-mapped compounds.

```{r}
geom.mean <- function(x) {
  # Returns the geometric mean
  exp(mean(log(x)))
}

ResolveAffinity <- function(df) {
  # Preferentially selects the affinity measure. If multiple meansurements
  # exist for the same compound-protein pair, the geometric mean is taken.
  for (measure in c('Kd', 'Ki', 'IC50')) {
    if (is.element(measure, df$measure)) {
      measure.df <- df[df$measure == measure, ]
      return.df <- data.frame(
        measure = measure,
        affinity_nM = round(geom.mean(measure.df$affinity_nM), 5),
        n_measures = nrow(measure.df),
        sources = paste(unique(na.omit(measure.df$source)), collapse=','),
        pubmeds = paste(unique(na.omit(measure.df$pubmed)), collapse=','))
      return(return.df)
    }
  }
}

# Create a single affinity measure for each compound-protein pair
collapse.df <- joined.df %>%
  dplyr::group_by(drugbank_id, bindingdb_id, uniprot, entrez_gene) %>%
  dplyr::do(ResolveAffinity(.)) %>%
  dplyr::ungroup()

collapse.df %>%
  readr::write_tsv('data/bindings-drugbank-collapsed.tsv')

# View a subset of the data.frame
collapse.df %>% dplyr::sample_n(200) %>% DT::datatable()
```

`r nrow(collapse.df)` compound--protein pairs were assayed.

```{r}
drugbank.df <- 'https://raw.githubusercontent.com/dhimmel/drugbank/3e87872db5fca5ac427ce27464ab945c0ceb4ec6/data/drugbank.tsv' %>%
  readr::read_tsv() %>%
  dplyr::mutate(drugbank_approved = as.integer(grepl('approved', groups))) %>%
  dplyr::transmute(drugbank_id, drugbank_name = name, drugbank_approved)  

entrez.df <- 'https://raw.githubusercontent.com/dhimmel/entrez-gene/5352b31e04ec136e99d25a0ba63e8867aa71b69f/data/genes-human.tsv' %>%
  readr::read_tsv() %>%
  dplyr::transmute(entrez_gene = GeneID, gene_symbol = Symbol)

gene.df <- collapse.df %>%
  dplyr::group_by(drugbank_id, entrez_gene) %>%
  dplyr::summarize(
    affinity_nM = min(affinity_nM),
    n_pairs = n(),
    sources = paste(unique(sources), collapse=','),
    pubmeds = paste(unique(pubmeds), collapse=',')
    ) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(drugbank.df) %>%
  dplyr::left_join(entrez.df)

gene.df %>%
  readr::write_tsv('data/bindings-drugbank-gene.tsv')

# View bindings for approved drugs
gene.df %>%
  dplyr::filter(affinity_nM <= 1000) %>%
  dplyr::filter(drugbank_approved == 1) %>%
  dplyr::select(drugbank_name, gene_symbol, affinity_nM, n_pairs) %>%
  DT::datatable()
```

`r nrow(gene.df)` drugbank--gene pairs have measured binding affinities.

### Interaction retention based on affinity threshold

```{r, fig.width=8}
exp.range <- -5:11

gene.df %>%
  ggplot(aes(x = affinity_nM)) +
  geom_histogram(alpha = 0.6) +
  scale_x_log10(
    breaks = scales::trans_breaks("log10", n=10, function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))) +
  theme_bw()

gene.df %>%
  ggplot(aes(x = affinity_nM)) +
  stat_ecdf() +
  scale_x_log10(
    breaks = scales::trans_breaks("log10", n=10, function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))) +
  theme_bw()
```

### Interactions per compound and per gene when restricting to micromolar or stronger affinities.

```{r, fig.width=8}
gene.df %>%
  dplyr::filter(affinity_nM <= 1000) %>%
  dplyr::group_by(drugbank_id) %>%
  dplyr::summarize(n_genes = n()) %>%
  ggplot(aes(x=n_genes)) +
  geom_histogram(alpha=0.6) +
  scale_x_log10(breaks=c(1:3, 5, 10, 20, 50, 100)) +
  xlab('Genes bound per compound') +
  theme_bw()

gene.df %>%
  dplyr::filter(affinity_nM <= 1000) %>%
  dplyr::group_by(entrez_gene) %>%
  dplyr::summarize(n_compounds = n()) %>%
  ggplot(aes(x=n_compounds)) +
  geom_histogram(alpha=0.6) +
  scale_x_log10(breaks=c(1:5, 7, 10, 15, 25, 50)) +
  xlab('Compounds binding per gene') +
  theme_bw()
```